---
title: "Epigenetic age acceleration, OCM micronutients, and metals"
subtitle: "Associations of prenatal one-carbon metabolism nutrients and metals with epigenetic aging biomarkers at birth and in childhood in a US cohort: Example code"
author: 
  - name: "Anne Bozack"
output:
  html_document:
    toc: true
    toc_float: true
---

## Libraries

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(DescTools)
library(Metrics)
library(MASS)
library(sandwich)
library(lmtest)
library(rms)
library(psych)
library(GPArotation)
library(knitr)
library(kableExtra)
```

## Load cord blood data
#### datCord: Dataset including the following variables
Exposure variables:

- fol_scale: mean centered and scaled first-trimester plasma folate concentration
- B12_scale: mean centered and scaled first-trimester plasma B12 concentration
- Cu_scale: mean centered and scaled first-trimester RBC Cu concentration
- Mg_scale: mean centered and scaled first-trimester RBC Mg concentration
- Mn_scale: mean centered and scaled first-trimester RBC Mn concentration
- Se_scale: mean centered and scaled first-trimester RBC Se concentration
- Zn_scale: mean centered and scaled first-trimester RBC Zn concentration
- As_scale: mean centered and scaled first-trimester RBC As concentration
- Ba_scale: mean centered and scaled first-trimester RBC Ba concentration
- Cd_scale: mean centered and scaled first-trimester RBC Cd concentration
- Cr_scale: mean centered and scaled first-trimester RBC Cr concentration
- Cs_scale: mean centered and scaled first-trimester RBC Cs concentration
- Hg_scale: mean centered and scaled first-trimester RBC Hg concentration
- Pb_scale: mean centered and scaled first-trimester RBC Pb concentration
- fol_scale_win: scaled and Winsorized first-trimester plasma folate concentration
- B12_scale_win: scaled and Winsorized first-trimester plasma B12 concentration
- Cu_scale_win: scaled and Winsorized first-trimester RBC Cu concentration
- Mg_scale_win: scaled and Winsorized first-trimester RBC Mg concentration
- Mn_scale_win: scaled and Winsorized first-trimester RBC Mn concentration
- Se_scale_win: scaled and Winsorized first-trimester RBC Se concentration
- Zn_scale_win: scaled and Winsorized first-trimester RBC Zn concentration
- As_scale_win: scaled and Winsorized first-trimester RBC As concentration
- Ba_scale_win: scaled and Winsorized first-trimester RBC Ba concentration
- Cd_scale_win: scaled and Winsorized first-trimester RBC Cd concentration
- Cr_scale_win: scaled and Winsorized first-trimester RBC Cr concentration
- Cs_scale_win: scaled and Winsorized first-trimester RBC Cs concentration
- Hg_scale_win: scaled and Winsorized first-trimester RBC Hg concentration
- Pb_scale_win: scaled and Winsorized first-trimester RBC Pb concentration

Covariates

- sex: child sex (0 = male; 1 = female)
- race_child2: child race and ethnicity (1 = White; 2 = Black; 3 = Hispanic; 4 = Asian; 5 = More than one race or other)
- mom_age: maternal age at enrollment
- mom_bmi: maternal prepregnancy BMI
- coll_grad: maternal education (0 = not collage graduate; 1 = college graduate)
- nullip: nulliparous
- gt70k: household income (0 = income <= 70k USD; 1 = income > 70k USD)
- smk_preg: maternal smoking (0 = never smoker; 1 = former smoker; 2 = smoking during pregnancy)

Cell type estimates

- Bcell: B cell proportion estimated in cord blood
- CD4T: CD4 T cell proportion estimated in cord blood
- Gran: granulocyte proportion estimated in cord blood
- Mono: monocyte proportion estimated in cord blood
- NK: natural killer cell proportion estimated in cord blood
- nRBC: nucleated RBC proportion estimated in cord blood

EAA variables

- GA_BohlinPred_EAA_wks: Bohlin epigenetic gestational age acceleration (weeks)
- AgeAccelerationResidual_cord_wks: Horvath epigenetic age acceleration (weeks)
- DNAmAgeSkinBloodClockAdjAge_cord_wks: skin & blood epigenetic age acceleration (weeks)

## Robust linear models for single metals and EAA
```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Function to run rlm models and store results
# dat: input data
# ageVar: EAA variable
# metals: vector of exposure variables
# covars: vector of covariates
# cellvars: vector of cell type estimates
rlmTab = function(dat, ageVar, metals, covars, cellvars){
	dfResults = data.frame(metals = metals, B_adjCell = NA, CI_lower_adjCell = NA, CI_upper_adjCell = NA, summary_adjCell = NA, p_adjCell = NA, N_adjCell = NA)
	for (i in 1:length(metals)){
		mod_adjCell = rlm(as.formula(paste(ageVar, ' ~ ', metals[i], ' + ', paste(covars,collapse=' + '), ' + ', paste(cellvars,collapse=' + '), sep = '')), data = dat, maxit = 30)
        	test = coeftest(mod_adjCell, vcovHC(mod_adjCell, type="HC0"))
			ci = coefci(x = mod_adjCell, level = 0.95, vcov. = vcovHC(mod_adjCell, type="HC0"))
            # fill in table with results
        	dfResults[i,2] = test[2,1] # B
			dfResults[i,3] = ci[2,1] # CI lower bound
			dfResults[i,4] = ci[2,2] # CI upper bound
    		dfResults[i,5] = paste0(round(test[2,1], 2), ' (', round(ci[2,1], 2), ', ', paste0(round(ci[2,2], 2), ')')) # summary of B and CI
        	dfResults[i,6] = round(test[2,4], 4) # p-value
			dfResults[i,7] =dim(mod_adjCell$model)[1] # N of data analyzed
	}
	return(dfResults)
}
```

### Robust linear models in cord blood 
```{r, warning=FALSE, message=FALSE, eval=FALSE}
metals = c('fol_scale', 'B12_scale', 'Cu_scale', 'Mg_scale', 'Mn_scale', 'Se_scale', 'Zn_scale', 'As_scale', 'Ba_scale', 'Cd_scale', 'Cr_scale', 'Cs_scale', 'Hg_scale', 'Pb_scale')
covars = c('sex', 'race', 'mom_age', 'mom_bmi', 'coll_grad', 'nullip', 'gt70k', 'smk_preg')
cellvars = c('Bcell', 'CD4T', 'CD8T', 'Gran', 'Mono', 'NK', 'nRBC')     

# Bohlin GAA
ageVar = 'GA_BohlinPred_EAA_wks'
rlm_Bohlin = rlmTab(datCord, ageVar, metals, covars, cellvars)

# Horvath EAA
ageVar = 'AgeAccelerationResidual_cord_wks'
rlm_Horvath_cord = rlmTab(datCord, ageVar, metals, covars, cellvars)

# Skin & blood EAA
ageVar = 'DNAmAgeSkinBloodClockAdjAge_cord_wks'
rlm_skinblood_cord = rlmTab(datCord, ageVar, metals, covars, cellvars)
```

### Linear associations with Bohlin EGAA
```{r, echo = F}
rlm_Bohlin %>% kable() %>% kable_styling(font_size = 12) 
```

### Linear associations with Horvath EAA
```{r, echo = F}
rlm_Horvath_cord %>% kable() %>% kable_styling(font_size = 12) 
```

### Linear associations with skin & blood EAA
```{r, echo = F}
rlm_skinblood_cord %>% kable() %>% kable_styling(font_size = 12) 
```

## Non-linear associations using restricted cubic splines 
```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Function to run RCS models and store results
# dat: input data
# ageVar: EAA variable
# metals: vector of exposure variables
# covars: vector of covariates and cell type estimates
rcsTab = function(dat, ageVar, metals, covars){
	dfResults = data.frame(metals = metals, B_1 = NA, B2 = NA, p_overall = NA, p_nonlinear = NA)
	for (i in 1:length(metals)){
        knots = as.character(quantile(dat[[metals[i]]], c(0.1, 0.5, 0.9)))
		fit_rcs = ols(as.formula(paste(ageVar, ' ~  rcs(', metals[i], ', c(', paste(knots,collapse = ', '), '))', ' + ', paste(covars,collapse=' + '), sep = '')), data = dat)
            dfResults[i,2] = fit_rcs$coefficients[2] # B1
            dfResults[i,3] = fit_rcs$coefficients[3] # B2
            dfResults[i,4] = anova(fit_rcs)[1,5] # p-value overall
            dfResults[i,5] = anova(fit_rcs)[2,5] # nonlinear p-value
	}
	return(dfResults)
}
```

### RCS in cord blood 
```{r, warning=FALSE, message=FALSE, eval=FALSE}
metals = c('fol_scale_win', 'B12_scale_win', 'Cu_scale_win', 'Mg_scale_win', 'Mn_scale_win', 'Se_scale_win', 'Zn_scale_win', 'As_scale_win', 'Ba_scale_win', 'Cd_scale_win', 'Cr_scale_win', 'Cs_scale_win', 'Hg_scale_win', 'Pb_scale_win')                                                                                                                                                                       
covars = c('sex', 'race', 'mom_age', 'mom_bmi', 'coll_grad', 'nullip', 'gt70k', 'smk_preg', 'Bcell', 'CD4T', 'CD8T', 'Gran', 'Mono', 'NK', 'nRBC')                                                                                                                                                                  

# Bohlin with Winsorization
ageVar = 'GA_BohlinPred_EAA_wks'                                                                                                                                                               
rcs_Bohlin = rcsTab(datCord, ageVar, metals, covars)

## Horvath with Winsorization
ageVar = 'AgeAccelerationResidual_cord_wks'                                                                                                                                                               
rcs_Horvath = rcsTab(datCord, ageVar, metals, covars)

## Skin & blood with Winsorization
ageVar = 'DNAmAgeSkinBloodClockAdjAge_cord_wks'                                                                                                                                                               
rcs_skinblood = rcsTab(datCord, ageVar, metals, covars)
```

### Nonlinear associations with Bohlin EGAA
```{r, echo = F}
rcs_Bohlin %>% kable() %>% kable_styling(font_size = 12) 
```

### Nonlinear associations with Horvath EAA
```{r, echo = F}
rcs_Horvath %>% kable() %>% kable_styling(font_size = 12) 
```

### Nonlinear associations with skin & blood EAA
```{r, echo = F}
rcs_skinblood %>% kable() %>% kable_styling(font_size = 12) 
```

## Explorabory factor analysis
```{r, warning=FALSE, message=FALSE, eval=FALSE}
# predictors
Z = datCord[,c('fol_scale', 'B12_scale', 'Cu_scale', 'Mg_scale', 'Mn_scale', 'Se_scale', 'Zn_scale', 'As_scale', 'Ba_scale', 'Cd_scale', 'Cr_scale', 'Cs_scale', 'Hg_scale', 'Pb_scale')]  
Z = as.matrix(Z)

# KMO test
KMO(Z)
# remove variables with MSA < 0.60 (B12, Ba, Cd, Cr)
Z <- Z[, KMO(Z)$MSAi>0.6]
KMO(Z)                                                                                                                                                                                                                                                                                                                       

# Bartlett test (compare correlation matrix to identify matrix)
cortest.bartlett(Z)
# p = 2.43331e-84

# parallel analysis scree plot
fa.parallel(Z, fa="fa")
# Parallel analysis suggests that the number of factors =  2  and the number of components =  NA 

# using 2 factors
fit2 <- fa(Z, nfactors = 2, rotate="oblimin", fm = 'ml')
summary(fit2)
# BIC =  -99.77

# add scores to datset
scores = factor.scores(Z, fit2)$scores
datCord = cbind(datCord, scores)
```

### Association between scores and EAA
```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Function to run rlm models and store results
# dat: input data
# ageVar: EAA variable
# factors: EFA scores
# covars: vector of covariates 
# cellvars: cell type estimates
rlmTab_efa = function(dat, ageVar, factors, covars, cellvars){
	dfResults = data.frame(factor = factors, B_adjCell = NA, CI_lower_adjCell = NA, CI_upper_adjCell = NA, summary_adjCell = NA, p_adjCell = NA, N_adjCell = NA)
    mod_adjCell = rlm(as.formula(paste(ageVar, ' ~ ', paste(factors,collapse=' + '), '+', paste(covars,collapse=' + '), ' + ', paste(cellvars,collapse=' + '), sep = '')), data = dat)
    test_adjCell = coeftest(mod_adjCell, vcovHC(mod_adjCell, type="HC0"))
	ci_adjCell = coefci(x = mod_adjCell, level = 0.95, vcov. = vcovHC(mod_adjCell, type="HC0"))
	for (i in 1:length(factors)){
        dfResults[i,2] = test_adjCell[i+1,1] # B
    	dfResults[i,3] = ci_adjCell[i+1,1] # lower CI
		dfResults[i,4] = ci_adjCell[i+1,2] # upper CI
		dfResults[i,5] = paste0(round(test_adjCell[i+1,1], 3), ' (', round(ci_adjCell[i+1,1], 3), ', ', paste0(round(ci_adjCell[i+1,2], 3), ')')) # summary of B and CI
        dfResults[i,6] = round(test_adjCell[i+1,4], 4) # p-value
		dfResults[i,7] = dim(mod_adjCell$model)[1] # N of data analyzed
    }
	return(dfResults)
}

factors = c('ML1', 'ML2')
covars = c('female_d', 'race_child2', 'age_mom_enroll_d', 'bmi_mom_prepreg_d', 'coll_grad', 'nullip', 'gt70k', 'smk_preg2')
cellvars = c('Bcell_GS_cb', 'CD4T_GS_cb', 'CD8T_GS_cb', 'Gran_GS_cb', 'Mono_GS_cb', 'NK_GS_cb', 'nRBC_GS_cb')     

# Bohlin EGAA
ageVar = 'GA_BohlinPred_EAA_wks'
rlm_Bohlin_efa = rlmTab_efa(datCord, ageVar, factors, covars, cellvars)

# Horvath EAA
ageVar = 'AgeAccelerationResidual_cord_wks'
rlm_Horvath_efa = rlmTab_efa(datCord, ageVar, factors, covars, cellvars)

# Skin & blood EAA
ageVar = 'DNAmAgeSkinBloodClockAdjAge_cord_wks'
rlm_skinblood_efa = rlmTab_efa(datCord, ageVar, factors, covars, cellvars)
```

### EFA associations with Bohlin EGAA
```{r, echo = F}
rlm_Bohlin_efa %>% kable() %>% kable_styling(font_size = 12) 
```

### EFA associations with Horvath EAA
```{r, echo = F}
rlm_Horvath_efa %>% kable() %>% kable_styling(font_size = 12) 
```

### EFA associations with skin & blood EAA
```{r, echo = F}
rlm_skinblood_efa %>% kable() %>% kable_styling(font_size = 12) 
```